Reproducible Research: Peer Assessment 2

Created by S. A. Batla on 20 November 2015

## Consequences of Severe Weather Events on Public Health and Economy in the United States  

### Synopsis

Storms and other severe weather events can cause both public health and 
economic problems for communities and municipalities. Many severe events can 
result in fatalities, injuries, and property damage, and preventing such 
outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric 
Administration (NOAA) storm database. This database tracks characteristics 
of major storms and weather events in the United States, including when and 
where they occur, as well as estimates of any fatalities, injuries, and 
property damage.

Basically, we want to answer explore and answer two questions:
1. Across the United States, which types of events (as indicated in the     
   EVTYPE variable) are most harmful with respect to population health? 
2. Across the United States, which types of events have the greatest 
   economic consequences?

### Basic settings & libraries

```{r}
echo = TRUE
library(R.utils)
library(ggplot2)
library(plyr)
library(dplyr)
library(gridExtra)
```

### Data Processing

```{r}
if (!"stormData" %in% ls()) {
stormData <- read.csv("data/StormData.csv", sep = ",") # this takes a bit
}

dim(stormData)
head(stormData)
```
There are 902,297 observations in the dataset

The description of the data states that events in the database start in the 
year 1950 and end in November 2011. It goes on to suggest that earlier years 
recorded in the database are not ideal for our analysis. More recent years 
should be considered. We plot a simple histogram to visualize this.

```{r}
stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, 
                                            format = "%m/%d/%Y %H:%M:%S"), "%Y"))
hist(stormData$year, breaks=30, col="blue")
```  

Based on the histogram above, we see that the number of events tracked starts to 
significantly increase around 1995. So, we subset the data from 1995 to 2011 to 
leverage more complete records for our analysis.

```{r}
stormData <- stormData[stormData$year >= 1995, ]
dim(stormData)
```
Now there are 681,500 observations in the dataset
In addition, not all variables are needed in the analysis, we will now remove
irrelevant variables from the dataset
Only the following fields are required:
STATE__,BGN_DATE,EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP

```{r}
stormData<-select(stormData,STATE,BGN_DATE,EVTYPE,FATALITIES,INJURIES,PROPDMG,
                   PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

Only 9 variables left in the dataset
Now, we inspect the exponents for property damage and crop damage, to see what 
data quality issues we might encounter there.

```{r}
summary(stormData$PROPDMGEXP)
summary(stormData$CROPDMGEXP)
```

Here we see some data quality issues with both PROPDMGEXP & CROPDMGEXP variables.
Namely, ?, +, 1, 0-8 and ?, 0, 2 for PROPDMGEXP & CROPDMGEXP, respectively.
We do not know what these mean. In addition, the units are mixed case in
some instances.

First, we uppercase the valid exponents
```{r}
stormData$PROPDMGEXP<-as.factor(toupper(stormData$PROPDMGEXP))
stormData$CROPDMGEXP<-as.factor(toupper(stormData$CROPDMGEXP))
```

Then, NA the unwanted exponents, add the correct multiplier for the valid 
exponents and total the damage for both Property and Crops - this adds another 
variable to the dataset
```{r}
pde<-as.character(stormData$PROPDMGEXP)
pde[pde %in% c("?","-","+","1","2","3","4","5","6","7","8")]<-NA
pde[pde %in% c("","0")]<-0
pde<-gsub("B", 10^9, pde)
pde<-gsub("K", 1000, pde)
pde<-gsub("H", 100, pde)
pde<-gsub("M", 10^6, pde)
stormData$PROPDMGEXP<-as.numeric(pde)
stormData$TOTALPROPDMG<-stormData$PROPDMGEXP*stormData$PROPDMG
head(stormData)

cde<-as.character(stormData$CROPDMGEXP)
cde[cde %in% c("?","2")]<-NA
cde[cde %in% c("","0")]<-0
cde<-gsub("B", 10^9, cde)
cde<-gsub("K", 1000, cde)
cde<-gsub("H", 100, cde)
cde<-gsub("M", 10^6, cde)
stormData$CROPDMGEXP<-as.numeric(cde)
stormData$TOTALCROPDMG<-stormData$CROPDMGEXP*stormData$CROPDMG
head(stormData)
```

As a last data tidying step, we total the **fatalities**, **injuries**, 
**property damages**, **crop damages**, and **total damages** by event type
```{r}
totalStormImpact<-ddply(stormData, .(EVTYPE), summarize,
                       totalFatalities=sum(FATALITIES),
                       totalInjuries=sum(INJURIES),
                       totalPropDamage=sum(TOTALPROPDMG),
                       totalCropDamage=sum(TOTALCROPDMG),
                       totalDamage=sum(totalPropDamage,totalCropDamage))
```
stormData is now a tidy dataset ready for analysis

### Results
#### Impact on Public Health
There are 985 levels of EVTYPE. Since we are only interested in weather events 
with the most impact on **fatalities** and **injuries**, we will consider only
the top 10 events.

```{r}
fatalities<-totalStormImpact[ order(-totalStormImpact[, 2]), ][1:10, ]
injuries<-totalStormImpact[ order(-totalStormImpact[,3]), ][1:10, ]

print(fatalities[,1:2], row.names=FALSE)
print(injuries[,1:3], row.names=FALSE)
```
Plot the Top 10 events with highest **Fatalities** and **Injuries**
  
```{r}
fatalitiesChart <- ggplot(data = fatalities, aes(y=totalFatalities, 
                                    x=reorder(EVTYPE, totalFatalities))) + 
  geom_bar(stat="identity", fill="red") +coord_flip() + 
  ylab("Fatalities") + 
  xlab("Weather Event") + 
  ggtitle("Top 10 Events by Total Fatalities")

injuriesChart <- ggplot(data = injuries, aes(y=totalInjuries, 
                                  x=reorder(EVTYPE, totalInjuries))) + 
  geom_bar(stat="identity", fill="blue")+ coord_flip() +
  ylab("Injuries") +
  xlab("Weather Event") + 
  ggtitle("Top 10 Events by Total Injuries")

grid.arrange(fatalitiesChart, injuriesChart, ncol=2)
```

According to the graph above, **excessive heat** is the leading cause of
**fatalities** and **tornados** are the leading cause of **injuries** - 
according to data collected by the NOAA from 1995-2011

#### Impact on Economy
There are 985 levels of EVTYPE. Since we are only interested in weather events 
with the most impact on **damages**, we will consider only the top 10 events.

```{r{}
propDamages<-totalStormImpact[ order(-totalStormImpact[, 4]), ][1:10, ]
cropDamages<-totalStormImpact[ order(-totalStormImpact[,5]), ][1:10, ]
totalDamages<-totalStormImpact[ order(-totalStormImpact[,6]), ][1:10, ]

print(propDamages[,1:4], row.names=FALSE)
print(cropDamages[,1:5], row.names=FALSE)
print(totalDamages[,1:6], row.names=FALSE)
```

Plot the Top 10 events with highest **Property Damage** and **Crop Damage**
```{r}
propDamageChart <- ggplot(data = propDamages, 
                          aes(y=totalPropDamage/10^9, 
                              x=reorder(EVTYPE, totalPropDamage))) + 
  geom_bar(stat="identity", fill="red") + coord_flip() + 
  ylab("Property Damages (US$ Billions)") + 
  xlab("Weather Event") + 
  ggtitle("Top 10 Events by Property Damages")

cropDamageChart <- ggplot(data = cropDamages, 
                          aes(y=totalCropDamage/10^9, 
                              x=reorder(EVTYPE, totalCropDamage))) + 
  geom_bar(stat="identity", fill="blue") + coord_flip() + 
  ylab("Total Damages (US$ Billions)") + 
  xlab("Weather Event") + 
  ggtitle("Top 10 Events by Crop Damages")

totalDamageChart <- ggplot(data = totalDamages, 
                          aes(y=totalDamage/10^9, 
                              x=reorder(EVTYPE, totalDamage))) + 
  geom_bar(stat="identity", fill="orange") + coord_flip() + 
  ylab("Total Damages (US$ Billions)") + 
  xlab("Weather Event") + 
  ggtitle("Top 10 Events by Total Damages")

grid.arrange(propDamageChart, cropDamageChart, totalDamageChart, ncol=2, nrow=2)
```

According to the graph above,  **flood** causes the highest financial 
impact on **property damage** and **drought** causes the highest financial 
impact on **crop damage** according to data collected by the NOAA from 1995-2011

### Conclusion  
Analysis conducted from data collected by the NOAA between 1995-2011 shows
that **excessive heat** and **tornado** events are most harmful to public
health and **flood**, **hurricane/typhoon**, and **storm surge** events have 
the greatest negative economic consequences - see third panel in last chart.

